Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)

### 2023-05-07 doc/manuscript/ESSDmanuscript.tex
update through end of section 5.1

\textbf{Abstract.} Forests are critical for climate change mitigation
and constitute a substantial portion of planned emissions reductions
under the 2015 Paris Agreement. Yet, the efficacy of greenhouse gas
mitigation planning and reporting is dependent upon the quality of
available emission factors data, including forest carbon (C) stocks and
changes therein. Tens of thousands of relevant forest C estimates have
been published, yet are not readily accesible to the practitioners
compiling national greenhouse gas inventories. Many of these data have,
however, been compiled in the Global Forest C database (ForC;
https://forc-db.github.io/) and stand to be of value to greenhouse gas
accounting if made available through the Emission Factor Database (EFDB)
of the International Panel on Climate Change (IPCC). Here, we develop
and document a process for semi-automated transfer of data from ForC
into the EFDB, assess the data available and transferred to date, and
provide recommendations for improving forest data collection, analysis,
and reporting to improve accounting of forest-sector greenhoouse gas
emissions and removals. We begin by reconciling terminology and mapping
ForC fields into EFDB. This process required some updates to the ForC
database structure, leading to the release of a new version of ForC
(v4.0; described here). As of May 06, 2023, ForC contained
\textasciitilde17204 independent records that would be relevant to EFDB,
1214 of which have been submitted to date. Among the data in ForC, there
is disproportionate representation of biomass (particularly aboveground)
stocks, with far fewer records for dead organic matter and soil C, and
relatively few or no records for net annual increments or C fluxes into
(gains) or out of (losses) the IPCC-defined C pools. Geographic
representation is also quite uneven, with the highest densities of
relevant records in temperate forests, and with relatively scant
representation of tropical forests in Africa and Asia. ForC represents a
diversity of stand ages, although records for young stands are primarily
limited to C stocks, as opposed to net increments of fluxes. This
distribution of records is generally reflected in the subset of records
that have been submitted to EFDB to date. In the future, forest C
estimates in EFDB can be improved through targeted research to fill
critical gaps, reporting of information required by IPCC, and continued
submission of data from scientific publications to the EFDB. Given that
climate change is rapidly impacting the world's forests, timely
reporting of recent estimates will be especially critical to accurate
forest C accounting.

@@

The Global Forest Carbon Database, ForC
(\url{https://forc-db.github.io/}), is the largest collection of
published estimates of forest C stocks, increments, and annual fluxes
\citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}. ForC
includes data ingested from individual publications and relevant
databases, including the Global Reforestation Opportunity Assessment
(GROA) database \citep[database doi:
10.5281/zenodo.3983644]{cook-patton_mapping_2020}, the global soil
respiration database
\citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
As of May 06, 2023, ForC contained 39855 records from 10589 plots in
1535 distinct geographical areas, along with records of stand age and
disturbance history. As such, ForC is positioned to improve forest C
accounting through the transfer of data to EFDB. The purpose of this
publication is to document that process and provide recommendations for
future improvements.

@@

We added a total of 15 new EFDB-relevant variables to the set of named
and defined variables (Fig. 1), counting each pair of variables with
units in C (ending in \texttt{\_C}) or organic matter (ending in
\texttt{\_OM}) as one. The majority of these were increment variables
(n=11), adding to only one previously defined increment variable
(aboveground biomass increment, \emph{delta.agb}). These are directly
related to C stocks as previously defined in ForC, with
``\emph{delta.}'' added in front of the variable name. Further, we added
variables capturing the belowground component of woody mortality
(\emph{woody.mortality\_root}) and the combined aboveground and
belowground components of woody mortality (\emph{woody.mortality}).
Although most of these variables lacked records in ForC as of May 06,
2023, their addition gave the structure such that records can be
populated over time. Finally, to provide better definition of the
previously existing variable \emph{organic.layer}, which has a nebulous
definition that reflects the varied definitions adopted by original
studies, we added two clearly defined variables: \emph{litter}
(relatively undecomposed plant material/ OL horizon), and
\emph{O.horizon} (entire O-horizon, including OL).

@@

To transfer complete, reviewed ForC records into EFDB, we created R
scripts to restructure ForC records and populate EFDB's bulk import form
(``EFDB bulk import.xlsx''). Criteria for data transfer were that (1)
records had been checked against the original study and determined to be
complete and correct (\texttt{EFDB.ready} field in
\texttt{ForC\_citations} table), (2) the original study presented values
in tables or text, as opposed to the values having been digitized from
graphs or calculated based on related variables
(\texttt{data.location.within.source} field in
\texttt{ForC\_measurements} table), and (3) the records had not
previously been sent to EFDB. Once converted into EFDB format, the
records were reviewed and then sent to the IPCC's Technical Support Unit
for inclusion in EFDB. Complete records needed to be reviewed by the
EFDB editorial board prior to posting in the database -- a process that
lags behind records transfer and had not yet been completed for all
records sent as of May 06, 2023.

@@

The mapping of ForC fields into EFDB fields is summarized in Appendix B,
with details documented in the public GitHub repository associated with
the project, IPCC-EFDB-integration repository within the ForC-db
organization (file \emph{ForC-EFDB\_mapping.csv} available at
\url{https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv}).

For the majority of fields, contents of the field in ForC was
transferred directly into an EFDB field, either as the only contents of
that field or as part of a composite record. For example, ten ForC
fields describing site location, climate, and edaphic properties all
mapped into the EFDB field \texttt{Region/Regional\ conditions}
(Appendix B). In cases where original studies did not present 95\%
confidence intervals (required by IPCC when available) but did present
information required to calculate these (standard error or n and
standard deviation), we calculated the 95\% confidence intervals and
populated the EFDB field with this information (noting the calculation
in the EFDB field \texttt{Comments\ from\ Data\ Provider}). For some
fields, simple conditional logic was used to populate EFDB fields based
on ForC records. For example, for stock variables presented in the
original publication in units of dry organic matter mass (as opposed to
C), several greenhouse gasses (CO\textsubscript{2}, CO,
CH\textsubscript{4}, NO, NO\textsubscript{2}, N\textsubscript{2}O) were
entered in the EFDB field indicating the greenhouse gases to which the
record could be pertinent (\texttt{Gases} field) because these values
could be used in calculations of greenhouse gas emissions from biomass
burning \citep{ipcc_2006_2006}; otherwise, the only pertinent greenhouse
gas would be CO\textsubscript{2}. There were two cases in which more
complex mapping was required: (1) mapping of C cycle variables (section
4.1.1) and (2) land classification (section 4.1.2).

@@

With input from the IPCC's Technical Support Unit, we reviewed the list
of ForC variables to identify those that were relevant to EFDB and to
appropriately map them into EFDB (Fig. 1). For each C pools (Table 1),
we identified variables representing organic matter or C stocks, stock
changes (a.k.a. ``net annual increments'' by IPCC, ``increments'' in
ForC), gains (a.k.a. ``gross annual increments'' by IPCC, ``fluxes'' in
ForC), and losses (``fluxes'' in ForC). As described in section 3.2, we
also defined 15 new EFDB-relevant variables that were not previously
represented in ForC. It it important to note that the correspondence of
ForC variables to IPCC criteria often depends upon measurement protocols
(``important sources of estimate variation'' in Table 1). For example,
ForC records of biomass and dead wood vary in the minimum stem diameter
censused, such that some records would match the IPCC criteria whereas
others would not. Information on minimum diameters censused and other
important sources of methodological variation are recorded as covariates
in ForC and mapped into the EFDB field \emph{Other Properties} (Appendix
B). Details on the mapping of ForC variables to EFDB -- including
associated covariates, IPCC pools (Table 1) and relevant equations
\citep{ipcc_2006_2006} -- are documented in the file
ForC\_variables\_mapping.csv in the IPCC-EFDB-integration repository
(\url{https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv}).

@@

As of May 06, 2023, ForC (v4.0) contained 32693 independent records
(39855 total), 17204 of which were for variables relevant to EFDB (Fig.
1). These records were distributed across all forested continents and
ecozones, with particularly high concentrations in
\emph{{[}ecozone(s){]}} and low concentrations in
\emph{{[}ecozone(s){]}} (Fig. 2). The most widely represented forest
type was \emph{{[}type{]}}, followed by \emph{{[}type{]}} and
\emph{{[}type{]}} (Fig. 3).

\begin{figure}
\includegraphics[width=15cm]{figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent} \caption{\textbf{Map of sites in ForC shaded by number of records relevant to (circles) and transferred to (triangles) EFDB.} Symbols are colored according to the number of records at each site. Underlying map shows FAO ecozones, which are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems.}\label{fig:fig_map}
\end{figure}

\newpage
\begin{figure}
\includegraphics[width=15cm]{figures_tables/Histogram_n_Relevant_and_Transferred_Records} \caption{\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types. For FAO ecozones (b), codes are as listed in the caption of Figure 2.}\label{fig:fig_histograms}
\end{figure}

ForC contained data for \textbf{\#\#} of the \textbf{\#\#} variables
relevant to EFDB (Table 2, Fig. 1). The records were very unevenly
distributed across variables. The most records colected were for the
Aboveground biomass variable(biomass\_ag).This variable

@@

As of May 06, 2023, we had reviewed or added 2034 EFDB-relevant records,
1214 records of which were sent to EFDB, and 73 of which have been
reviewed, accepted, and posted (Figs. 2-3, Table 2).
\emph{{[}DETAILS{]}}

@@

Table B1: \textbf{Mapping of ForC fields to EFDB.} See footnotes at end
of table (STILL NEED TO BE PROPERLY INSERTED).
\begingroup\fontsize{8}{10}\selectfont

>>>>

\textbf{Abstract.} Forests are critical for climate change mitigation
and constitute a substantial portion of planned emissions reductions
under the 2015 Paris Agreement. Yet, the efficacy of greenhouse gas
mitigation planning and reporting is dependent upon the quality of
available emission factors data, including forest carbon (C) stocks and
changes therein. Tens of thousands of relevant forest C estimates have
been published, yet are not readily accesible to the practitioners
compiling national greenhouse gas inventories. Many of these data have,
however, been compiled in the Global Forest C database (ForC;
https://forc-db.github.io/) and stand to be of value to greenhouse gas
accounting if made available through the Emission Factor Database (EFDB)
of the International Panel on Climate Change (IPCC). Here, we develop
and document a process for semi-automated transfer of data from ForC
into the EFDB, assess the data available and transferred to date, and
provide recommendations for improving forest data collection, analysis,
and reporting to improve accounting of forest-sector greenhoouse gas
emissions and removals. We begin by reconciling terminology and mapping
ForC fields into EFDB. This process required some updates to the ForC
database structure, leading to the release of a new version of ForC
(v4.0; described here). As of May 07, 2023, ForC contained
\textasciitilde17204 independent records that would be relevant to EFDB,
1214 of which have been submitted to date. Among the data in ForC, there
is disproportionate representation of biomass (particularly aboveground)
stocks, with far fewer records for dead organic matter and soil C, and
relatively few or no records for net annual increments or C fluxes into
(gains) or out of (losses) the IPCC-defined C pools. Geographic
representation is also quite uneven, with the highest densities of
relevant records in temperate forests, and with relatively scant
representation of tropical forests in Africa and Asia. ForC represents a
diversity of stand ages, although records for young stands are primarily
limited to C stocks, as opposed to net increments of fluxes. This
distribution of records is generally reflected in the subset of records
that have been submitted to EFDB to date. In the future, forest C
estimates in EFDB can be improved through targeted research to fill
critical gaps, reporting of information required by IPCC, and continued
submission of data from scientific publications to the EFDB. Given that
climate change is rapidly impacting the world's forests, timely
reporting of recent estimates will be especially critical to accurate
forest C accounting.

@@

The Global Forest Carbon Database, ForC
(\url{https://forc-db.github.io/}), is the largest collection of
published estimates of forest C stocks, increments, and annual fluxes
\citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}. ForC
includes data ingested from individual publications and relevant
databases, including the Global Reforestation Opportunity Assessment
(GROA) database \citep[database doi:
10.5281/zenodo.3983644]{cook-patton_mapping_2020}, the global soil
respiration database
\citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
As of May 07, 2023, ForC contained 39855 records from 10589 plots in
1535 distinct geographical areas, along with records of stand age and
disturbance history. As such, ForC is positioned to improve forest C
accounting through the transfer of data to EFDB. The purpose of this
publication is to document that process and provide recommendations for
future improvements.

@@

We added a total of 15 new EFDB-relevant variables to the set of named
and defined variables (Fig. 1), counting each pair of variables with
units in C (ending in \texttt{\_C}) or organic matter (ending in
\texttt{\_OM}) as one. The majority of these were increment variables
(n=11), adding to only one previously defined increment variable
(aboveground biomass increment, \emph{delta.agb}). These are directly
related to C stocks as previously defined in ForC, with
``\emph{delta.}'' added in front of the variable name. Further, we added
variables capturing the belowground component of woody mortality
(\emph{woody.mortality\_root}) and the combined aboveground and
belowground components of woody mortality (\emph{woody.mortality}).
Although most of these variables lacked records in ForC as of May 07,
2023, their addition gave the structure such that records can be
populated over time. Finally, to provide better definition of the
previously existing variable \emph{organic.layer}, which has a nebulous
definition that reflects the varied definitions adopted by original
studies, we added two clearly defined variables: \emph{litter}
(relatively undecomposed plant material/ OL horizon), and
\emph{O.horizon} (entire O-horizon, including OL).

@@

To transfer complete, reviewed ForC records into EFDB, we created R
scripts to restructure ForC records and populate EFDB's bulk import form
(``EFDB bulk import.xlsx''). Criteria for data transfer were that (1)
records had been checked against the original study and determined to be
complete and correct, and as originally presented, (2) the original
study presented values in tables or text, as opposed to the values
having been digitized from graphs or calculated based on related
variables, and (3) the records had not previously been sent to EFDB.
Once converted into EFDB format, the records were reviewed and then sent
to the IPCC's Technical Support Unit for inclusion in EFDB. Complete
records needed to be reviewed by the EFDB editorial board prior to
posting in the database -- a process that lags behind records transfer
and had not yet been completed for all records sent as of May 07, 2023.

@@

The mapping of ForC fields into EFDB fields is summarized in Appendix B.
For the majority of fields, contents of the field in ForC was
transferred directly into an EFDB field, either as the only contents of
that field or as part of a composite record. For example, ten ForC
fields describing site location, climate, and edaphic properties all
mapped into the EFDB field \emph{Region/Regional conditions} (Appendix
B). In cases where original studies did not present 95\% confidence
intervals (required by IPCC when available) but did present information
required to calculate these (standard error or n and standard
deviation), we calculated the 95\% confidence intervals and populated
the EFDB field with this information (noting the calculation in the EFDB
field \emph{Comments from Data Provider}). For some fields, simple
conditional logic was used to populate EFDB fields based on ForC
records. For example, for stock variables presented in the original
publication in units of dry organic matter mass (as opposed to C),
several greenhouse gasses (CO\textsubscript{2}, CO, CH\textsubscript{4},
NO, NO\textsubscript{2}, N\textsubscript{2}O) were entered in the EFDB
field indicating the greenhouse gases to which the record could be
pertinent (\emph{Gases} field) because these values could be used in
calculations of greenhouse gas emissions from biomass burning
\citep{ipcc_2006_2006}; otherwise, the only pertinent greenhouse gas
would be CO\textsubscript{2}. There were two cases in which more complex
mapping was required: (1) mapping of C cycle variables (section 4.1.1)
and (2) land classification (section 4.1.2).

@@

With input from the IPCC's Technical Support Unit, we reviewed the list
of ForC variables to identify those that were relevant to EFDB and to
appropriately map them into EFDB (Fig. 1). For each C pool (Table 1), we
identified variables representing organic matter or C stocks, stock
changes (a.k.a. ``net annual increments'' by IPCC, ``increments'' in
ForC), gains (a.k.a. ``gross annual increments'' by IPCC, ``fluxes'' in
ForC), and losses (``fluxes'' in ForC). As described in section 3.2, we
also defined 15 new EFDB-relevant variables that were not previously
represented in ForC. It it important to note that the correspondence of
ForC variables to IPCC criteria often depends upon measurement protocols
(``important sources of estimate variation'' in Table 1). For example,
ForC records of biomass and dead wood vary in the minimum stem diameter
censused, such that some records would match the IPCC criteria whereas
others would not. Information on minimum diameters censused and other
important sources of methodological variation are recorded as covariates
in ForC and mapped into the EFDB field \emph{Other Properties} (Appendix
B). Details on the mapping of ForC variables to EFDB -- including
associated covariates, IPCC pools (Table 1) and relevant equations
\citep{ipcc_2006_2006} -- are documented in the file
ForC\_variables\_mapping.csv in the GitHub repository associated with
this publication IPCC-EFDB-integration repository in ForC-db
organization (\url{https://github.com/forc-db/IPCC-EFDB-integration}).

@@

As of May 07, 2023, ForC (v4.0) contained 32693 independent records
(39855 total), 17204 of which were for the the 36 variables relevant to
EFDB (Fig. 1). These records were distributed across all forested
continents and ecozones (Fig. 2). The largest number of records came
from Asia, followed by North America, South America, and Europe, with
relatively few records from Africa, Australia, and Oceania (Fig. 3c).
Categorized by FAO ecozone, the greatest numbers of records came from
subtropical humid forests, temperate mountain systems, and tropical rain
forests, each with \textgreater2,000 independent records (Fig. 3b).
Boreal coniferous forests, temperate continental forests, subtropical
mountain systems, and tropical moist deciduous forests had
\textgreater1,000 independent records each, while other ecozones all had
\textless1,000 records. The most widely represented forest type was
needleleaf evergreen, followed by broadleaf deciduous and broadleaf
evergreen (Fig. 3a). In terms of stand age, the most represented age
class was 20-100 years, followed by \textless20 years and then
\textgreater100 years (Fig. 3d).

\begin{figure}
\includegraphics[width=15cm]{figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent} \caption{\textbf{Map of sites in ForC shaded by number of independent records relevant to (circles) and transferred to (triangles) EFDB.} Symbols are colored according to the number of records at each site. Underlying map shows FAO ecozones, which are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems.}\label{fig:fig_map}
\end{figure}

\newpage
\begin{figure}
\includegraphics[width=15cm]{figures_tables/Histogram_n_Relevant_and_Transferred_Records} \caption{\textbf{Histograms of number of independent records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types. For FAO ecozones (b), codes are as listed in the caption of Figure 2.}\label{fig:fig_histograms}
\end{figure}

ForC contained records for 22 of the 36 variables relevant to EFDB
(Table 2, Fig. 1). The records were very unevenly distributed across
variables. The variable with most records was aboveground biomass,
representing 45\% of all independent records relevant to EFDB, and
aboveground biomass components (woody biomass or foliage) representing
an additional 5\%. A total of 30\% of relevant records were for root
biomass (including fine and coarse root components), while 5\% described
total biomass. The non-living pools were less represented, with 5\% of
relevant were for dead wood (including standing and fallen components),
0.4\% for litter, and 2.3\% for soil carbon.

Increment and flux variables were poorly represented (Table 2). The
increment variable with most records was the aboveground biomass
increment, representing 0.7\% of all independent records relevant to
EFDB. The only other relevant increment variable with any records was
the O horizon (litter) increment, with just 4 records. Relevant flux
variable records (n=974) were limited to the biomass pools (aboveground,
belowground, or total) and together constituted 6\% of ForC's
independent records relevant to EFDB.

@@

As of May 07, 2023, we had reviewed or added 2034 EFDB-relevant records,
1214 records of which were sent to EFDB, and 73 of which have been
reviewed, accepted, and posted (Figs. 2-3, Table 2).
\emph{{[}DETAILS{]}}

@@

Table B1: \textbf{Mapping of ForC fields to EFDB.} Details documented in
the public GitHub repository associated with the project,
IPCC-EFDB-integration repository within the ForC-db organization (file
\emph{ForC-EFDB\_mapping.csv} available at
\url{https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv}).
See footnotes at end of table (STILL NEED TO BE PROPERLY INSERTED).
\begingroup\fontsize{8}{10}\selectfont
